EDIpack2: interoperable Lanczos-based solver for generic quantum impurity problems

### 2025-05-05 03_edipack2.tex
Fix a few compilation issues > mostly related to missing or duplicated bib entries

\paragraph{{\bf Remark}} As extensively discussed in
\onlinecite{amaricci2022}, the inclusion of local phonons with
truncated dimensions ultimately amounts to a sequential application of
the procedures defined for the electronic part, i.e. one per phonon
mode. This includes a linear scaling of the dimensions with $N_{ph}$
and of course largely limit the available degrees of freedom to
describe electronic states. In the rest of the paper we focus
specifically on the electronic part of the quantum impurity problem,
referring to the phonons in specific sections when their presence
introduces non-trivial modifications.

@@

The {\bf normal} case deals with independent conservation of the total occupation $N$
and the total magnetization $S_z$ or, equivalently, the total number of electrons with spin up $N_\up$ and down
$N_\dw$. Optionally, we consider the case in which the symmetry
applies separately for each orbital and spin,
i.e. $\vec{N_\s}=[N^1_\s,\dots,N^{N_\a}_\s]$. This special case has been
extensively discussed in \onlinecite{Amaricci2022} so it won't be covered in
this work.
%
The {\bf superc} case deals with the conservation of the total
magnetization only, so that total charge may not be conserved. This
case includes a description of $s$-wave superconductivity, featuring
intra- and inter-orbital components.
%
Finally, the {\bf nonsu2} case consider the conservation of the total
charge, whereas the spin symmetry group is not fully
conserved. Although there many possible realization of this scenario,
this particularly applies to the presence of local spin-orbit coupling
$\vec{L}\cdot\vec{S}$~\cite{something}, the emergence of in-plane spin ordering~\cite{KM} or
in-plane spin-triplet exciton condensation~\cite{ExcitonPRB,Amaricci,Blason}.  
%
From a computational point of view the construction of a symmetry
sector corresponds to the determination of a injective map
$\MM:\SS_{\vec{Q}}\rightarrow \FF$ relating the states $\ket{i}$
belonging to the sector $\SS_{\vec{Q}}$ to the states $\ket{I}$ in the
Fock space. Operatively, the map corresponds to an integer rank-1
array of dimension $D_\SS$ which is the {\it dimension} of the
sector. 
The table~\ref{TabSector} summarizes the properties of the symmetry
sectors.

@@

The {\bf normal} case requires a brief remark. Because of the
independent conservation of $N_\up$ and $N_\dw$, the local Hilbert
space and the electronic Fock can be factorizes as, respectively,
$\HH=\HH_\up\otimes\HH_\dw$, $\FF_e = \FF_{e\up}\otimes \FF_{e\dw}$.  
Accordingly any Fock state is written as $\ket{\vec{n}_\up}\ket{\vec{n}_\dw}$, the symmetry sector can be written as  $\SS_{\vec{Q}} = \SS_{N_\up}\otimes
\SS_{N_\dw}$ and, correspondingly, the sector map splits in two
mutually exclusive ones $\MM = \MM_\up
\otimes \MM_\dw$.
Thus, each state $\ket{i}=\ket{i_\up}\ket{i_\dw}$ of the
sector is labelled by two integers $[i_\up,i_\dw]$, 
$i_\sigma=1,\dots,D_{\SS_\sigma}$ such that $i=i_\up + i_\dw
D_{\SS_\dw}$. The map $\MM$ connects any such state to a Fock state
$\ket{I}=\ket{I_\up}\ket{I_\dw}$ labelled by two integers
$[I_\up,I_\dw]$ as $I=I_\up +   I_\dw 2^{N_s}$. For a thorough
discussion about Fock basis organization in this case see
\onlinecite{amaricci2022}.

The presence of a symmetry induces a factorization of the Fock space,
in turn inducing a block diagonal form to the Hamiltonian matrix.
Each block, labelled by the quantum numbers $\vec{Q}$, has dimension
$D_{\SS(\vec{Q})}$. The sector Hamiltonian matrix $H_\SS$ is represented in the
basis $\ket{i}\in\SS_{\vec{Q}}$ as a sparse matrix. In the {\bf
  normal} case $H_\SS$ takes a particularly symmetric form thanks to
the product structure of the sector and its map~\cite{amaricci2022}.
The analysis of the spectrum is then reconducted to the inspection of
the Hamiltonian in each symmetry sector. Should particular constraint
holds the search can be limited to only particular sector, further
reducing the computational cost.

@@

This operations easily takes over more than 80\% of the computation
time, so its optimization is a crucial aspect in any diagonalization
algorithms. In \onlinecite{amaricci2022} we covered in full details
some distrubuted memory massively parallel algorithms to address this
specific issue.
Here, using the same naming convention introduced in \cite{amaricci}, 
we briefly outline the main strategy for the implementation of the MVP
according to the value of {\tt ed\_mode}.

@@

\paragraph{{\bf superc/nonsu2}}
In this two cases the symmetry enforce a specific relation between $\up$ and
$\dw$ configurations ultimately suppressing the benefits arising from
Fock space factorization.
The electronic part of the Hamiltonian matrix is usually organized in this structure:
$$
H_\SS = H_{imp} + H_{int} + H_{bath} + H_{imp-bath} 
$$
where, with almost obvious meaning of the labels, $H_{imp}$ encodes
the non-interacting part of the impurity Hamiltonian determined by
$h^0_{\a\b\s\s'}$, $H_{int}$ the interaction term, $H_{bath}$ the
effective bath terms related to $h^p_{\a\b\s\s'}$ and finally
$H_{imp-bath}$ describes the coupling between impurity and bath
proportional to $V^p_{\a\b\s\s'}$. 
In an MPI setup the first three terms contains elements local in the
memory of each node. These are stored separately from the non-local
ones in any {\tt sparse\_matrix} representing $H_\SS$. 
The MVP is carried out using the {\tt MPI\_AllGatherV} algorithm
which, as discussed in \onlinecite{amaricci2022}, requires the
reconstruction of the distributed input array any node producing a
overflow in the MPI communication.

@@

\subsection{Reduced impurity density matrix}\label{sSecRDM}
This version of the \NAME library features the calculation of the
impurity Reduced Density Matrix (iRDM), enabling the analysis of
entanglement properties of the quantum impurity for any value of {\tt
  ed\_mode}.
To simplify the discussion in this section we take the zero
temperature limit and consider the presence of a non-degenerate
groundstate $\ket{\psi}$ populating the {\tt state\_list}.
The generalization to the presence of degenerate states or the finite
temperature regime are straightforward.
The groundstate belongs to a given symmetry sector $\SS$ and it
represented in the Fock basis as:  $\ket{\psi} = \sum_I a_I \ket{I}$. 
The density matrix $\rho$ is defined as:
$$
\rho = \ket{\psi}\bra{\psi} = \sum_{nm=1}^{4^{N_s}} a^*_ma_n \ket{n}\bra{m} =\sum_{nm=1}^{4^{N_s}} \rho_{nm} \ket{n}\bra{m} 
$$
The iRDM is obtained from this by tracing out the bath degrees of
freedom: $\rho^{i}=Tr_{b}\rho$. The summation appearing in the
previous expressions becomes quickly unmaneagable with the system
size. To overcome this issue we implemented a fast algorithm
leveraging on symmetry sector and map sparsity.
Using quantum number conservation the summation extrema get reducedto
the size $D_S$ of the sector $S$ to which the groundstate (or in
general any eigenstate) belongs. 
Next we analyze the groundstate expansion on the Fock states, splitting the bit representation of any state $\ket{I}$ into
their spin-dependent impurity and bath parts as $\ket{I} = \ket{
  i_\uparrow\,  b_\uparrow\, i_\downarrow \, b_\downarrow}$. Each bit
set span a space of dimension $D^\a_{imp}$ or $D^\s_{bath}$. 
Finally, we store into the {\tt sparse\_map} ${\cal S}$, for every impurity configuration 
$i_\s =1,\dots,D^\s_{imp}$ (key) the corresponding bath states $j\in[1,\dots,D^\s_{bath}]$
(values) which contribute to that key, as detailed in
\secu{CodeSparseMap}.
We indicate with ${\cal D}^\s_{i}$ the number of values for the key $i_\s$.
The fast summation algorithm for the evaluation of the iRDM takes on a
different form for the two cases {\bf normal} and {\bf superc/nonsu2},
as outlined below.

@@

{\bf REMARK TODO}
Notes on the norm function: a) the definition for conventional
matrices, ;b) the definition for the replica/general case where you
can choose between different norm defition.

>>>>

\paragraph{{\bf Remark}} As extensively discussed in
\onlinecite{Amaricci2022}, the inclusion of local phonons with
truncated dimensions ultimately amounts to a sequential application of
the procedures defined for the electronic part, i.e. one per phonon
mode. This includes a linear scaling of the dimensions with $N_{ph}$
and of course largely limit the available degrees of freedom to
describe electronic states. In the rest of the paper we focus
specifically on the electronic part of the quantum impurity problem,
referring to the phonons in specific sections when their presence
introduces non-trivial modifications.

@@

The {\bf normal} case deals with independent conservation of the total occupation $N$
and the total magnetization $S_z$ or, equivalently, the total number of electrons with spin up $N_\up$ and down
$N_\dw$. Optionally, we consider the case in which the symmetry
applies separately for each orbital and spin,
i.e. $\vec{N_\s}=[N^1_\s,\dots,N^{N_\a}_\s]$. This special case has been
extensively discussed in \onlinecite{Amaricci2022} so it won't be covered in
this work.
%
The {\bf superc} case deals with the conservation of the total
magnetization only, so that total charge may not be conserved. This
case includes a description of $s$-wave superconductivity, featuring
intra- and inter-orbital components.
%
Finally, the {\bf nonsu2} case consider the conservation of the total
charge, whereas the spin symmetry group is not fully
conserved. Although there many possible realization of this scenario,
this particularly applies to the presence of local spin-orbit coupling
$\vec{L}\cdot\vec{S}$~\cite{something}, the emergence of in-plane spin ordering~\cite{BellomiaKMH} or
in-plane spin-triplet exciton condensation~\cite{Amaricci2023_excitons,Blason}.  
%
From a computational point of view the construction of a symmetry
sector corresponds to the determination of a injective map
$\MM:\SS_{\vec{Q}}\rightarrow \FF$ relating the states $\ket{i}$
belonging to the sector $\SS_{\vec{Q}}$ to the states $\ket{I}$ in the
Fock space. Operatively, the map corresponds to an integer rank-1
array of dimension $D_\SS$ which is the {\it dimension} of the
sector. 
The table~\ref{TabSector} summarizes the properties of the symmetry
sectors.

@@

The {\bf normal} case requires a brief remark. Because of the
independent conservation of $N_\up$ and $N_\dw$, the local Hilbert
space and the electronic Fock can be factorizes as, respectively,
$\HH=\HH_\up\otimes\HH_\dw$, $\FF_e = \FF_{e\up}\otimes \FF_{e\dw}$.  
Accordingly any Fock state is written as $\ket{\vec{n}_\up}\ket{\vec{n}_\dw}$, the symmetry sector can be written as  $\SS_{\vec{Q}} = \SS_{N_\up}\otimes
\SS_{N_\dw}$ and, correspondingly, the sector map splits in two
mutually exclusive ones $\MM = \MM_\up
\otimes \MM_\dw$.
Thus, each state $\ket{i}=\ket{i_\up}\ket{i_\dw}$ of the
sector is labelled by two integers $[i_\up,i_\dw]$, 
$i_\sigma=1,\dots,D_{\SS_\sigma}$ such that $i=i_\up + i_\dw
D_{\SS_\dw}$. The map $\MM$ connects any such state to a Fock state
$\ket{I}=\ket{I_\up}\ket{I_\dw}$ labelled by two integers
$[I_\up,I_\dw]$ as $I=I_\up +   I_\dw 2^{N_s}$. For a thorough
discussion about Fock basis organization in this case see
\onlinecite{Amaricci2022}.

The presence of a symmetry induces a factorization of the Fock space,
in turn inducing a block diagonal form to the Hamiltonian matrix.
Each block, labelled by the quantum numbers $\vec{Q}$, has dimension
$D_{\SS(\vec{Q})}$. The sector Hamiltonian matrix $H_\SS$ is represented in the
basis $\ket{i}\in\SS_{\vec{Q}}$ as a sparse matrix. In the {\bf
  normal} case $H_\SS$ takes a particularly symmetric form thanks to
the product structure of the sector and its map~\cite{Amaricci2022}.
The analysis of the spectrum is then reconducted to the inspection of
the Hamiltonian in each symmetry sector. Should particular constraint
holds the search can be limited to only particular sector, further
reducing the computational cost.

@@

This operations easily takes over more than 80\% of the computation
time, so its optimization is a crucial aspect in any diagonalization
algorithms. In \onlinecite{Amaricci2022} we covered in full details
some distrubuted memory massively parallel algorithms to address this
specific issue.
Here, using the same naming convention introduced in \cite{Amaricci2022}, 
we briefly outline the main strategy for the implementation of the MVP
according to the value of {\tt ed\_mode}.

@@

\paragraph{{\bf superc/nonsu2}}
In this two cases the symmetry enforce a specific relation between $\up$ and
$\dw$ configurations ultimately suppressing the benefits arising from
Fock space factorization.
The electronic part of the Hamiltonian matrix is usually organized in this structure:
$$
H_\SS = H_{imp} + H_{int} + H_{bath} + H_{imp-bath} 
$$
where, with almost obvious meaning of the labels, $H_{imp}$ encodes
the non-interacting part of the impurity Hamiltonian determined by
$h^0_{\a\b\s\s'}$, $H_{int}$ the interaction term, $H_{bath}$ the
effective bath terms related to $h^p_{\a\b\s\s'}$ and finally
$H_{imp-bath}$ describes the coupling between impurity and bath
proportional to $V^p_{\a\b\s\s'}$. 
In an MPI setup the first three terms contains elements local in the
memory of each node. These are stored separately from the non-local
ones in any {\tt sparse\_matrix} representing $H_\SS$. 
The MVP is carried out using the {\tt MPI\_AllGatherV} algorithm
which, as discussed in \onlinecite{Amaricci2022}, requires the
reconstruction of the distributed input array any node producing a
overflow in the MPI communication.

@@

%GBtodo placeholder: revise this subsection
\subsection{Reduced impurity density matrix}\label{sSecRDM}
This version of the \NAME library features the calculation of the
impurity Reduced Density Matrix (iRDM), extending the algorithm introduced in \onlinecite{BellomiaPhD} so to enable the analysis of
entanglement properties of the quantum impurity for any value of {\tt
  ed\_mode}.
To simplify the discussion in this section we take the zero
temperature limit and consider the presence of a non-degenerate
groundstate $\ket{\psi}$ populating the {\tt state\_list}.
The generalization to the presence of degenerate states or the finite
temperature regime are straightforward.
The groundstate belongs to a given symmetry sector $\SS$ and it
represented in the Fock basis as:  $\ket{\psi} = \sum_I a_I \ket{I}$. 
The density matrix $\rho$ is defined as:
$$
\rho = \ket{\psi}\bra{\psi} = \sum_{nm=1}^{4^{N_s}} a^*_ma_n \ket{n}\bra{m} =\sum_{nm=1}^{4^{N_s}} \rho_{nm} \ket{n}\bra{m} 
$$
The iRDM is obtained from this by tracing out the bath degrees of
freedom: $\rho^{i}=Tr_{b}\rho$. The summation appearing in the
previous expressions becomes quickly unmaneagable with the system
size. To overcome this issue we implemented a fast algorithm
leveraging on symmetry sector and map sparsity.
Using quantum number conservation the summation extrema get reducedto
the size $D_S$ of the sector $S$ to which the groundstate (or in
general any eigenstate) belongs. 
Next we analyze the groundstate expansion on the Fock states, splitting the bit representation of any state $\ket{I}$ into
their spin-dependent impurity and bath parts as $\ket{I} = \ket{
  i_\uparrow\,  b_\uparrow\, i_\downarrow \, b_\downarrow}$. Each bit
set span a space of dimension $D^\a_{imp}$ or $D^\s_{bath}$. 
Finally, we store into the {\tt sparse\_map} ${\cal S}$, for every impurity configuration 
$i_\s =1,\dots,D^\s_{imp}$ (key) the corresponding bath states $j\in[1,\dots,D^\s_{bath}]$
(values) which contribute to that key, as detailed in
\secu{CodeSparseMap}.
We indicate with ${\cal D}^\s_{i}$ the number of values for the key $i_\s$.
The fast summation algorithm for the evaluation of the iRDM takes on a
different form for the two cases {\bf normal} and {\bf superc/nonsu2},
as outlined below.

@@

{\bf REMARK TODO}
Notes on the norm function: a) the definition for conventional
matrices, ;b) the definition for the replica/general case where you
can choose between different norm defition.
{\color{red} I (GB) can take care of this, as I introduced the "elemental" norm in the code.}
%GBtodo placeholder
